# Dummy Encoding {#sec-dummy-encoding}

We have some categorical variable and we want to turn it into numerical values, one of the most common ways of going about it it so create **dummy variables**. Dummy variables are variables that only takes the values 0 and 1 to indicate the absence or presence of the levels in a categorical variable. This is nicely shown with an example.

Consider this short categorical variable of animals, we observe there are 3 unique values "cat", "dog", and "horse".

```{r}
#| echo: false
c("dog", "cat", "horse", "dog", "cat")
```

Which just this knowledge we can create the corresponding dummy variables. There should be 3 columns one for each of the levels

```{r}
#| echo: false
dummy <- matrix(0L, nrow = 5, ncol = 3)
colnames(dummy) <- c("cat", "dog", "horse")
dummy[1, 2] <- 1L
dummy[2, 1] <- 1L
dummy[3, 3] <- 1L
dummy[4, 2] <- 1L
dummy[5, 1] <- 1L
dummy
```

From this we have a couple of observations. Firstly the length of each of these variables are equal to the length of the original categorical variable. The number of columns corresponds to the number of levels. And lastly the sum of all the values on each row equals 1 since all the rows contains one 1 and the remaining 0s.

talk about how this is a trained method

Explain one hot vs dummy

TODO add Diagram

## Pros and Cons

### Pros

### Cons

## R Examples

# Python Examples

TODO
