---
pagetitle: "Feature Engineering A-Z | Periodic Splines"
aliases: 
  - circular-splines.html
---

# Periodic Splines {#sec-periodic-splines}

::: {style="visibility: hidden; height: 0px;"}
## Periodic Splines
:::

This chapter expands the idea we saw in the [last chapter](periodic-trig.qmd) with the use of [splines](numeric-splines.qmd). These splines allow for different shapes of activation than we saw with trigonomic functions.

We will be using the same toy data set and see if we can improve on it.

```{r}
#| label: toy_data
#| echo: false
#| message: false
library(tidyverse)
library(recipes)
library(patchwork)
set.seed(1234)

toy_data <- tibble(
  predictor = c(1, 1000, sample(1:1000, 300))
) |>
  mutate(
    target = pmax(sin(predictor / 365 * pi * 2), 0)^4 +
      rnorm(n(), sd = 0.1) -
      0.5
  ) |>
  arrange(target)
```

```{r}
#| label: fig-spline-predictor-outcome
#| echo: false
#| message: false
#| fig-cap: |
#|   Strong periodic signal every 365 values along predictor.
#| fig-alt: |
#|   Scatter chart. Predictor (0-2000) on x-axis, target on y-axis. Data shows
#|   clear periodic pattern with period 365: target stays near -0.5 most of
#|   the time, with sharp peaks rising to about 0.5 occurring every 365 units.
toy_data |>
  ggplot(aes(predictor, target)) +
  geom_point() +
  theme_minimal()
```

This data has a very specific shape, 
and we will see if we can approcimate it with our splines.

First we fit a number of spline terms to our data using default arguments.

```{r}
#| label: fig-periodic-spline-defaults
#| echo: false
#| message: false
#| fig-cap: |
#|   Strong periodic signal every 365 values along predictor below, spline values above.
#| fig-alt: |
#|   Two stacked charts. Bottom: scatter plot showing periodic target data
#|   with peaks every 365 units. Top: default periodic B-spline basis
#|   functions - multiple overlapping curves that span the entire predictor
#|   range but don't match the data's period. The splines are too wide and
#|   not aligned with the 365-unit cycle.
p1 <- recipe(~., data = toy_data) |>
  step_spline_b(
    predictor,
    options = list(periodic = TRUE),
    complete_set = TRUE,
    keep_original_cols = TRUE
  ) |>
  prep() |>
  bake(NULL) |>
  pivot_longer(-c(predictor, target)) |>
  ggplot(aes(predictor, value, color = name)) +
  geom_line() +
  theme_minimal() +
  guides(color = "none") +
  labs(x = NULL, y = "activation")

p2 <- toy_data |>
  ggplot(aes(predictor, target)) +
  geom_point() +
  theme_minimal()

p1 / p2
```

While it produces some fine splines they are neither well fitting or periodic.
Let us make spline periodic and try to approcimate the period.

```{r}
#| label: fig-periodic-spline-period
#| echo: false
#| message: false
#| fig-cap: |
#|   Strong periodic signal every 365 values along predictor below, spline values above.
#| fig-alt: |
#|   Two stacked charts. Bottom: scatter plot showing periodic target data.
#|   Top: periodic B-spline basis with Boundary.knots set to [0, 365]. The
#|   spline curves now repeat every 365 units, matching the data's period.
#|   Multiple bell-shaped curves tile the space, each activating for a
#|   different phase of the cycle.
p1 <- recipe(~., data = toy_data) |>
  step_spline_b(
    predictor,
    options = list(periodic = TRUE, Boundary.knots = c(0, 365)),
    complete_set = TRUE,
    keep_original_cols = TRUE
  ) |>
  prep() |>
  bake(NULL) |>
  pivot_longer(-c(predictor, target)) |>
  ggplot(aes(predictor, value, color = name)) +
  geom_line() +
  theme_minimal() +
  guides(color = "none") +
  labs(x = NULL, y = "activation")

p2 <- toy_data |>
  ggplot(aes(predictor, target)) +
  geom_point() +
  theme_minimal()

p1 / p2
```

We already see that something good is happening,
The width of each bump is related to the number of degrees of freedom we have,
lowering this value creates more wider bumps.

```{r}
#| label: fig-periodic-spline-calibrated
#| echo: false
#| message: false
#| fig-cap: |
#|   Strong periodic signal every 365 values along predictor below, spline values above.
#| fig-alt: |
#|   Two stacked charts. Bottom: scatter plot showing periodic target peaks.
#|   Top: calibrated periodic B-spline with deg_free=7 and shifted boundary
#|   knots. One spline term highlighted in purple aligns almost perfectly
#|   with the peak locations in the data below. Other terms shown in gray
#|   cover different phases of the cycle.
p1 <- recipe(~., data = toy_data) |>
  step_spline_b(
    predictor,
    degree = 3,
    deg_free = 7,
    options = list(periodic = TRUE, Boundary.knots = c(0, 365) + 50),
    complete_set = TRUE,
    keep_original_cols = TRUE
  ) |>
  prep() |>
  bake(NULL) |>
  pivot_longer(-c(predictor, target)) |>
  ggplot(aes(predictor, value, color = name)) +
  geom_line() +
  theme_minimal() +
  guides(color = "none") +
  labs(x = NULL, y = "activation") +
  scale_color_manual(values = c(rep("gray80", 6), "#643BA0"))

p2 <- toy_data |>
  ggplot(aes(predictor, target)) +
  geom_point() +
  theme_minimal()

p1 / p2
```

Now we got some pretty good traction.
Pulling out the well performing spline term,
we can translate it a bit to show how well it overlaps with our signal.

```{r}
#| label: fig-periodic-spline-fit
#| echo: false
#| message: false
#| fig-cap: |
#|   Strong periodic signal every 365 values along predictor.
#| fig-alt: |
#|   Scatter chart with fitted spline overlay. Points show periodic target
#|   data with peaks every 365 units. A purple curve (scaled and shifted
#|   spline term) traces through the peaks.
recipe(~., data = toy_data) |>
  step_spline_b(
    predictor,
    degree = 3,
    deg_free = 7,
    options = list(periodic = TRUE, Boundary.knots = c(0, 365) + 50),
    complete_set = TRUE,
    keep_original_cols = TRUE
  ) |>
  prep() |>
  bake(NULL) |>
  select(predictor, target, spline = predictor_7) |>
  mutate(spline = spline * 1.7 - 0.5) |>
  ggplot(aes(predictor, target)) +
  geom_point() +
  geom_line(
    aes(predictor, spline),
    color = "#643BA0",
    linewidth = 1,
    alpha = 0.75
  ) +
  theme_minimal()
```

::: {.callout-note}
While one spline term is highlighted here,
It is important to note that the coverage of the splines makes sure that any signal is captured.
:::

There are obviously some signals that can't be captured using splines.
Compared to sine curves they are much more flexible, 
with a number of different kinds,
each with some room for customization.
Any purely periodic signal can be captured in the [next chapter](periodic-indicators.qmd).

## Pros and Cons

### Pros

- More flexible than sine curves
- Fairly interpretable

### Cons

- Requires that you know the period
- Will create some unnecessary features
- Can't capture all types of signal

## R Examples

We will be using the animalshelter data set for this.

```{r}
#| label: show-data
library(recipes)
library(animalshelter)

longbeach |>
  select(outcome_type, intake_date)
```

There are two steps in the recipes package that support periodic splines.
Those are `step_spline_b()` and `step_spline_nonnegative()`,
used for B-splines and Non-negative splines (also called M-Splines) respectively.

These functions have 2 main arguments controlling the spline itself,
and 2 main arguments controlling its periodic behavior.

`deg_free` and `degree` controls the spline,
changing the number of spline terms that are created,
and the degrees of the piecewise polynomials respectively.
The defaults for these functions tend to be a good starting point.
To make these steps periodic,
we need to set `periodic = TRUE` in `options`.
Lastly, we can control the period and its shift with `Boundary.knots` in `options`.
I find the easiest way to set this like this: `c(0, period) + shift`.

```{r}
#| label: step_spline_b
spline_rec <- recipe(outcome_type ~ intake_date, data = longbeach) |>
  step_mutate(intake_date = as.integer(intake_date)) |>
  step_spline_b(
    intake_date,
    options = list(periodic = TRUE, Boundary.knots = c(0, 365) + 50)
  )

spline_rec |>
  prep() |>
  bake(new_data = NULL)
```

::: callout-caution
# TODO

Find dataset where the predictor is naturally numeric.
:::


## Python Examples
