---
pagetitle: "Feature Engineering A-Z | ISOMAP"
---

# ISOMAP {#sec-too-many-isomap}

::: {style="visibility: hidden; height: 0px;"}
## ISOMAP
:::

Isometric mapping (ISOMAP) is a non-linear dimensionality reduction method,
much like [UMAP](too-many-umap.qmd).
It is another method that uses distances between points to produce graphs of neighboring points. 
What diffirentiate this method from the others is that it uses **geodesic distances** as opposed to straight-line distances.
where the geodesic distance is the sum of edge weights along the shortest path between two points.

The eigenvectors of the deodesic distance metric are then used to represent the new coordinates.

A very high-level description of the Isomap algorithm is given below:

1. Find the neighbors for each point
1. Construct the neighborhood graph, using Euclidean distance as edge length
1. Calculate the shortest path between each pair of points
1. Use Multidimensional scaling to compute a lower-dimensional embedding

## Pros and Cons

### Pros

Captures non-linear effects
Captures long-range structure, not just local structure
No parameters to set other than neighbors

### Cons

Computationally expensive
Assumes a single connected manifold

## R Examples


```{r}
#| label: ames
#| echo: false
#| message: false
library(tidymodels)
data("ames")
```

We will be using the `ames` data set for these examples.

```{r}
#| label: show-data
library(recipes)
library(modeldata)

ames_num <- ames |>
  select(where(is.numeric))
```

{embed} provides `step_isomap()`, which is the standard way to perform isomap.

```{r}
#| label: step_isomap
isomap_rec <- recipe(~ ., data = ames_num) |>
  step_normalize(all_numeric_predictors()) |>
  step_isomap(all_numeric_predictors(), num_terms = 5)

isomap_rec |>
  prep() |>
  bake(new_data = NULL) |>
  glimpse()
```

## Python Examples

WIP
